---
title: "Assessment"
output: html_document
date: "2025-10-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

#### Has the COVID-19 pandemic influenced antidepressant prescribing patterns patterns during the Freshers' season (September-October) across Scottish health boards?


#Join the data boards

#I loaded september to october data boards from 2017 - 2023 to represent the freshers season then I merged the health boards 
#I summarise total antidepressant prescriptions per Freshers year and plotted the graph to see the trend 
#Looked at pre coviid , during covid and after covid trend to see if there has been any impact or association
```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(here) # directory stucture
library(gt) # tables
library(janitor) # cleaning data
```


```{r, message = FALSE, warning = FALSE}
freshers_all <- read_csv(here("freshers_all.csv"))
```

```{r, message = FALSE, warning = FALSE}
freshers_clean_all <- freshers_all %>% 
  select(hbt2014,bnf_item_code,bnf_item_description,number_of_paid_items,paid_quantity,paid_date_month)
```

```{r}
freshers_clean_all2 <- freshers_clean_all %>% 
  mutate(paid_date_month = as.character(paid_date_month),year=as.numeric(substr(paid_date_month,1,4)))
# summarise total prescriptions by year during freshers season "201709" so you need first four digits

```

```{r}
freshers_summary <- freshers_clean_all2 %>% 
group_by(year) %>% 
  summarise(total_items=sum(number_of_paid_items,na.rm = TRUE))

#GROUPS BY YEAR
library("ggplot2")
ggplot(freshers_summary,aes(x=year,y=total_items))+
  geom_line(linewidth=0.7,colour="blue")+
  geom_point(size=2)+
  scale_x_continuous(breaks =2017:2023) +
  #to show the labels 2017 and 2013
  labs(title="Antidepressant Prescriptions During Freshers Season",x="Year",y="Total Antidepressant Prescriptions") +
  theme_minimal()
```

# I also will separate the different seasons and look at them more into detail. (pre-covid, covid, post-covid)
# I also want to show each region separately by health boards 



