---
title: "Assessment"
output: html_document
date: "2025-10-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```

## R Markdown

#### Has the COVID-19 pandemic influenced antidepressant prescribing patterns during the winteseason (September-October) across Scottish health boards?

## Winter season

## scotishcensusgov

#Join the data boards

I loaded september to october data boards from 2017 - 2023 to represent the freshers season then I merged the health boards #I summarise total antidepressant prescriptions per Freshers year and plotted the graph to see the trend #Looked at pre coviid , during covid and after covid trend to see if there has been any impact or association

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(here) # directory stucture
library(gt) # tables
library(janitor) # cleaning data
library(ggplot2) # plotting graph
library(sf) # to read in map data 
library(readxl) # to read in map data
library(plotly) # to make interactive
library(viridis)
library(sf)

```

loading a large amount of data in a shorter time period by downloading and using the mapdfr function (data from 2017-2023)

```{r message=FALSE, warning=FALSE}
files <- list.files(here("data", "winter_data"), pattern = "csv")
winter_data <- files %>% 
  map_dfr(~read_csv(here("data", "winter_data", .))) %>% 
clean_names()
```

clean up data and filter for the sections you want

```{r message=FALSE, warning=FALSE}
filtered_winter_data <- winter_data %>% 
filter(str_starts(bnf_item_code,"0403")) %>%  #antidepressant code is 0403
  mutate(year = as.numeric(substr(paid_date_month,1,4)), month = as.numeric(substr(paid_date_month,5,6))) %>% #separates the date into years and month so that i can group winter sections
  mutate(winter_year=case_when(month == 12 ~ year + 1, 
month %in% c(1,2) ~ year) )#makes a new column to group the winter years 

filtered_winter_data <- filtered_winter_data %>% 
  unite("healthboards",hbt2014,hbt,sep = "_")#so some of my data healthboard codes were under the name hbt_2014 AND another was hbt so i had to merge the column so all the healthboard columns fall under one

  filtered_winter_data$healthboards <- gsub("[NA]","",filtered_winter_data$healthboards) 
    filtered_winter_data$healthboards <-
      gsub("_","",filtered_winter_data$healthboards)#had to remove some NA characters and '_' characters

```

Graph 1

```{r message=FALSE, warning=FALSE}
winter_years_data <- filtered_winter_data %>% 
  group_by(winter_year) %>% 
  summarise(total_items=sum(number_of_paid_items,na.rm = TRUE))

plot <- ggplot(winter_years_data,aes(x=winter_year,y=total_items)) +
  geom_line(linewidth=0.7,colour = "blue") +
  geom_point(size=2)+
  scale_x_continuous(breaks=2017:2023) +
  labs(title="Antidepressant Prescriptions During Winter Season",x="Year",y="Total Antidepressant Prescriptions") +
  theme_minimal()
  
ggplotly(plot)
print(plot)

#write a code talking about the zoomed in changes and reference why you dudnt go from 0

```

```{r}
#ROUGH
population <- readxl::read_excel(here("data","population.xlsx"), skip=10) %>% 
  clean_names() %>% 
  group_by(x2,all_people) %>% 
summarise () %>% 
  filter(!is.na(all_people))

population <- population %>% 
  rename(h_bname = x2)

filtered2_winter_data <- filtered_winter_data %>% 
  group_by(healthboards,bnf_item_code,paid_quantity,winter_year,gp_practice) %>% summarise(total_paid = sum(paid_quantity, na.rm =TRUE))

```

```{r}
SIMD <- readxl::read_excel(here("data","SIMD.xlsx")) %>% 
  clean_names() # loading excel data 
filtered_SIMD <- SIMD %>% 
  group_by(simd2020v2_quintile,h_bcode,h_bname) %>% 
  summarise()

filtered_SIMD <- filtered_SIMD %>% 
  rename(healthboards = h_bcode)

Overall_SIMD_winter <- filtered2_winter_data %>% 
  full_join(filtered_SIMD,by = "healthboards")
'relationship = "many-to-many"'

Overall_SIMD_population_winter <- Overall_SIMD_winter %>% 
  full_join(population,by="h_bname")

```

```{r}

antidepressant_per_head <- Overall_SIMD_population_winter %>% 
  group_by(healthboards,h_bname,winter_year) %>% 
  summarise(quantity_per_head = sum(paid_quantity)/mean(all_people))

### map 
NHS_healthboards <- st_read(here( "data", "Week6_NHS_HealthBoards_2019.shp")) %>% 
clean_names() %>% 
  rename(h_bname = hb_name)

# Join spatial data with falls_admissions_75_summary
mapped_data <- antidepressant_per_head %>%
  full_join(NHS_healthboards,by="h_bname") %>% 
  st_as_sf()


#CLAUDE : 
  library(ggiraph)

plot_map <- mapped_data %>% 
  ggplot() + 
  geom_sf_interactive(  # Changed from geom_sf
    aes(fill = quantity_per_head,
        tooltip = paste0(h_bname, 
                        "\nWinter Year: ", winter_year,
                        "\nQuantity per Head: ", round(quantity_per_head, 2))),
    colour = "white", 
    size = 0.1
  ) + 
  scale_fill_distiller(palette = "Blues", direction = 1, 
                       name = "Items per Head") +
  labs(
    title = "Antidepressant Prescriptions per Head",
    subtitle = "By Health Board and Winter Year"
  ) +
  facet_wrap(~ winter_year) +
  theme_void() +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 10)
  )

interactive_map <- girafe(ggobj = plot_map)  # Changed from ggplotly
interactive_map
```


CHAT

```{r}
#summarise totals per winter year per practice
winter_year_summary <- Overall_SIMD_population_winter %>% 
  group_by (healthboards, gp_practice, winter_year) %>% 
  summarise(total_paid = sum(paid_quantity, na.rm = TRUE))
# average across winter years
winter_year_average <- winter_year_summary %>% 
  group_by(healthboards,gp_practice) %>%
  summarise(avg_paid_over_winters = mean(total_paid, na.rm = TRUE))
#SIMD
winter_with_simd <- winter_year_average %>% 
  full_join(filtered_SIMD, by="healthboards") %>% 
  filter(!is.na(simd2020v2_quintile))
```

CHAT2

```{r}
box <- ggplot(winter_with_simd,
       aes(x=factor(simd2020v2_quintile),
           y=avg_paid_over_winters,
           fill=factor(simd2020v2_quintile))) +
  geom_boxplot(outlier.shape = 21,
               outlier.size = 1.5,
               outlier.stroke = 0.5,
               linewidth = 0.8,
               colour = "black",
               alpha = 0.7) +
  scale_y_continuous(labels = scales::label_number())+
  scale_fill_viridis(discrete=TRUE, alpha=0.9) +
  geom_jitter(color='red',size=0.4, alpha=0.4) +
 theme(
  panel.background = element_rect(fill = "lightblue",
                                colour = "lightblue",
                                size = 0.5)) +
  labs(
    title="Average Winter Antidepressant Prescriptions per Practice by SIMD Quintile",
    x="SIMD Quintile (1 = Most Deprived)",
    y="Avg prescriptions per practice (winter seasons)")
ggplotly(box)
```

PERCENTAGE CHANGE
```{r}

SIMD_winter_summary <- winter_year_summary %>% 
  full_join(filtered_SIMD,by = "healthboards")
'relationship = "many-to-many"'

SIMD_winter_final <- SIMD_winter_summary %>% 
  mutate(period =ifelse( winter_year < 2020, "pre","post"))  %>% 
group_by(simd2020v2_quintile,h_bname, period) %>%
  summarise(mean_total = mean(total_paid, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(
    names_from = period,
    values_from = mean_total
  ) %>%
  mutate(
    pct_change = ((post - pre) / pre) * 100) %>% 
 mutate(h_bname = reorder(h_bname, simd2020v2_quintile)) 

SIMD_winter_final <- SIMD_winter_final %>% 
   filter(!is.na(pct_change))
```

chat3
```{r}
library(ggplot2)
library(dplyr)

lollipop <- SIMD_winter_final %>%
  ggplot(aes(
    y = reorder(h_bname, -simd2020v2_quintile),   # or reorder(h_bname, pct_change)
    x = pct_change
  )) +
  
  # stick from 0 → % change
  geom_segment(
    aes(x = 0, xend = pct_change, yend = h_bname),
    linewidth = 1,
    colour = "grey40"
  ) +
  
  # lollipop dot
  geom_point(
    aes(x = pct_change),
    size = 2,
    colour = ifelse(SIMD_winter_final$pct_change > 0, "blue", "red")
  ) +
  
  # percent labels on dots
  geom_text(
    aes(label = paste0(round(pct_change, 1), "%")),
    hjust = ifelse(SIMD_winter_final$pct_change > 0, -0.2, 1.2),
    size = 3
  ) +
  
  scale_x_continuous(labels = function(x) paste0(x, "%")) +

  labs(
    title = "Percentage Change in Antidepressant Prescriptions",
    subtitle = "Lollipop plot (COVID Post vs Pre), ordered by SIMD",
    x = "Percentage Change",
    y = "Health Board"
  ) +
  
  theme_minimal(base_size = 13)

ggplotly(lollipop)
```
 



questions : not sure the best way to displa my original data ? github - how to get rid of the signs 1- overall national trend (use original graph) 2- i want to show the variation between different regions using healthboards 3- link it to deprevation and look at prescriptions per person 4- can i do a map that shows pre covid and post covid side by side would that count as one 5- voilin plot across differet SIMDs to compare smaller unit of data - gp practice (postcode that links to SIMD) PATCHWORK - MAPS TOGETHETE reference line to show the split between precovid, covid and postcovid

every dot is a gp practice - gp practice - dataset - adressess (assessment prep) voilin plot if messy add transperency open data use quintiles for voilin plot do a code that says if not installed install and load packages

percentage increase


overall trend
map antidepressant prescribing per head by healthboard facet by winter year?
boxplot by SIMD 
dumbell plot / lollipop graph - percentage change 

